---
title: "Caries_Biomarkers"
author: "Chiranjit Mukherjee"
date: "11/3/2020"
output: html_document
  keep_md=TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library("randomForest")
library(plyr) # for the "arrange" function
library(rfUtilities) # to test model significance
library(caret) # to get leave-one-out cross-validation accuracies and also contains the nearZeroVar function
library(e1071)
library(ggplot2)
library(vegan)
```


```{r}
otu_table <- readRDS("data/v13_counts_hd.rds")
metadata <- readRDS("data/meta_hd.rds")
```



```{r}

# Convert to relative abundance
otu_table_pct <- decostand(otu_table, method = "total")

# Compute NMDS
set.seed(12345); capture.output(otu_table_pct.mds <- metaMDS(otu_table_pct, trymax = 200, autotransform = F, wascores = F))

# Permanova
set.seed(12345); otu_table_pct.perm <- adonis(formula = otu_table_pct ~ metadata$status) # R2 0.25679, p < 0.001***

# Making dataframe for plotting
otu_table_pct.mds.df <- data.frame(scores(otu_table_pct.mds, display = 'sites'))
otu_table_pct.mds.df$status <- metadata$status[match(row.names(otu_table_pct.mds.df), row.names(metadata))]
```


```{r}
# Plot NMDS
ggplot(otu_table_pct.mds.df, aes(x=NMDS1, y=NMDS2)) + stat_ellipse(alpha=0.8, aes(color=status), show.legend = F, lwd=0.2) + 
  geom_point(alpha=0.9, aes(fill = status), size=3, color="black", pch=21, stroke=0.2) + scale_fill_manual(values=c("firebrick4", "forestgreen")) + scale_color_manual(values=c("firebrick4", "forestgreen")) + 
  labs(title = "Beta Diversity Comparison", subtitle = "Relative Abundance", fill="Subject status") +
  geom_text(aes(label=rownames(otu_table_pct.mds.df)), size=2) + 
  annotate("text", x = (min(otu_table_pct.mds.df$NMDS1) + max(otu_table_pct.mds.df$NMDS1))/2, y = max(otu_table_pct.mds.df$NMDS2)+0.1, 
           label = paste("p <", otu_table_pct.perm$aov.tab$`Pr(>F)`[1], "(PERMANOVA)", sep=" ")) + theme_classic() +
  theme(plot.title = element_text(size=15, face="bold", hjust=0.5), plot.subtitle = element_text(size=10, hjust=0.5), 
        axis.title = element_text(size=10, face="bold"), axis.text = element_text(size=8, face="bold"), legend.position="bottom", 
        axis.line = element_line(size = 0.3),legend.title = element_text(size=13, face="bold"), legend.text = element_text(size = 11))
ggsave(file = "figs/otu_table_pct_mds.pdf", width = 10, height = 6, units = "in")
```




```{r}
# Removing EG_204_d as an outlier
otu_table_filt <- otu_table[row.names(otu_table) != "EG_204_d", ]
dim(otu_table_filt)
metadata_filt <- metadata[row.names(metadata) != "EG_204_d", ]
```

```{r}
# Convert to relative abundance
otu_table_filt_pct <- decostand(otu_table_filt, method = "total")

# Compute NMDS
set.seed(12345); capture.output(otu_table_filt_pct.mds <- metaMDS(otu_table_filt_pct, trymax = 200, autotransform = F, wascores = F))

# Permanova
set.seed(12345); otu_table_filt_pct.perm <- adonis(formula = otu_table_filt_pct ~ metadata_filt$status) # R2 0.25679, p < 0.001***
otu_table_filt_pct.perm # R2: 0.27237, p < 0.001 ***

# Making dataframe for plotting
otu_table_filt_pct.mds.df <- data.frame(scores(otu_table_filt_pct.mds, display = 'sites'))
otu_table_filt_pct.mds.df$status <- metadata_filt$status[match(row.names(otu_table_filt_pct.mds.df), row.names(metadata_filt))]
```

```{r}

# Plot NMDS
ggplot(otu_table_filt_pct.mds.df, aes(x=NMDS1, y=NMDS2)) + stat_ellipse(alpha=0.8, aes(color=status), show.legend = F, lwd=0.2) + 
  geom_point(alpha=0.9, aes(fill = status), size=3, color="black", pch=21, stroke=0.2) + scale_fill_manual(values=c("firebrick4", "forestgreen")) +
  scale_color_manual(values=c("firebrick4", "forestgreen")) + 
  labs(title = "Beta Diversity Comparison", subtitle = "Relative Abundance", fill="Subject status") +
  #geom_text(aes(label=rownames(otu_table_filt_pct.mds.df)), size=3) + 
  annotate("text", x = (min(otu_table_filt_pct.mds.df$NMDS1) + max(otu_table_filt_pct.mds.df$NMDS1))/2, y = max(otu_table_filt_pct.mds.df$NMDS2)+0.1, 
           label = paste("p <", otu_table_filt_pct.perm$aov.tab$`Pr(>F)`[1], "(PERMANOVA)", sep=" ")) + theme_classic() +
  theme(plot.title = element_text(size=15, face="bold", hjust=0.5), plot.subtitle = element_text(size=10, hjust=0.5), 
        axis.title = element_text(size=10, face="bold"), axis.text = element_text(size=8, face="bold"), legend.position="bottom", 
        axis.line = element_line(size = 0.3), legend.title = element_text(size=13, face="bold"), legend.text = element_text(size = 11))
ggsave(file = "figs/otu_table_filt_pct_mds.pdf", width = 10, height = 6, units = "in")
```


#### Filter out rare OTUs 
Keeping only those OTUs which have less than 30 (75% of total number of samples) times 0 values in the dataset
```{r}
otu_table_filt_rare_removed <- otu_table_filt[, colSums(otu_table_filt == 0) < 0.75*nrow(otu_table_filt), ]
dim(otu_table_filt_rare_removed) #  [1] 63 x 166 (166 species remain, of the 343)
```

```{r}
# Convert to relative abundance
otu_table_filt_rare_removed_pct <- decostand(otu_table_filt_rare_removed, method = "total")

# Compute NMDS
set.seed(12345); capture.output(otu_table_filt_rare_removed_pct.mds <- metaMDS(otu_table_filt_rare_removed_pct, trymax = 200, autotransform = F, wascores = F))

# Permanova
set.seed(12345); otu_table_filt_rare_removed_pct.perm <- adonis(formula = otu_table_filt_rare_removed_pct ~ metadata_filt$status) # R2 0.25679, p < 0.001***
otu_table_filt_rare_removed_pct.perm # R2: 0.28585, p < 0.001 ***

# Making dataframe for plotting
otu_table_filt_rare_removed_pct.mds.df <- data.frame(scores(otu_table_filt_rare_removed_pct.mds, display = 'sites'))
otu_table_filt_rare_removed_pct.mds.df$status <- metadata_filt$status[match(row.names(otu_table_filt_rare_removed_pct.mds.df), row.names(metadata_filt))]
```

```{r}

# Plot NMDS
ggplot(otu_table_filt_rare_removed_pct.mds.df, aes(x=NMDS1, y=NMDS2)) + stat_ellipse(alpha=0.8, aes(color=status), show.legend = F, lwd=0.2) + 
  geom_point(alpha=0.9, aes(fill = status), size=3, color="black", pch=21, stroke=0.2) + scale_fill_manual(values=c("firebrick4", "forestgreen")) +
  scale_color_manual(values=c("firebrick4", "forestgreen")) + 
  #geom_text(aes(label=rownames(otu_table_filt_rare_removed_pct.mds.df)), size=3) + 
  annotate("text", x = (min(otu_table_filt_rare_removed_pct.mds.df$NMDS1) + max(otu_table_filt_rare_removed_pct.mds.df$NMDS1))/2, 
           y = max(otu_table_filt_rare_removed_pct.mds.df$NMDS2)+0.1, 
           label = paste("p <", otu_table_filt_rare_removed_pct.perm$aov.tab$`Pr(>F)`[1], "(PERMANOVA)", sep=" ")) + theme_classic() +
  labs(title = "Beta Diversity Comparison", subtitle = "Relative Abundance", fill="Subject status") +
  theme(plot.title = element_text(size=15, face="bold", hjust=0.5), plot.subtitle = element_text(size=10, hjust=0.5), 
        axis.title = element_text(size=10, face="bold"), axis.text = element_text(size=8, face="bold"), legend.position="bottom", 
        axis.line = element_line(size = 0.3), legend.title = element_text(size=13, face="bold"), legend.text = element_text(size = 11))
ggsave(file = "figs/otu_table_filt_rare_removed_pct_mds.pdf", width = 10, height = 6, units = "in")
```






```{r}
set.seed(123); RF_status_classify <- randomForest(x=otu_table_filt_rare_removed_pct, y=metadata_filt$status,
                                                 ntree=501, importance=TRUE, proximities=TRUE )
RF_status_classify
```




```{r}
oob.error.data <- data.frame(
  Trees=rep(1:nrow(RF_status_classify$err.rate), times=3),
  Type=rep(c("OOB", "Caries", "Healthy"), each=nrow(RF_status_classify$err.rate)),
  Error=c(RF_status_classify$err.rate[,"OOB"], 
    RF_status_classify$err.rate[,"Caries"], 
    RF_status_classify$err.rate[,"Healthy"]))

ggplot(data=oob.error.data, aes(x=Trees, y=Error)) + geom_line(aes(color=Type))
# ggsave("oob_error_rate_500_trees.pdf")
```


### Permutation Testing Model Performance

```{r, eval=FALSE}
#Evaluate is set to false to reduce Knit time
RF_status_classify_sig <- rf.significance( x=RF_status_classify, xdata=otu_table_filt_rare_removed_pct,
                                          nperm=1000 , ntree=501 )
RF_status_classify_sig
```

### Accuracy Estimated by Cross-validation

```{r}
fit_control <- trainControl(method = "LOOCV")    
RF_status_classify_loocv <- train(x=otu_table_filt_rare_removed_pct, y=metadata_filt$status,
                                  method="rf", ntree=501 , tuneGrid=data.frame(mtry=25), trControl=fit_control)
RF_status_classify_loocv

```


## Identifying Important Features
```{r}
RF_status_classify_imp <- as.data.frame(RF_status_classify$importance)
RF_status_classify_imp$features <- rownames(RF_status_classify_imp)
RF_status_classify_imp_25 <- RF_status_classify_imp[order(RF_status_classify_imp$MeanDecreaseAccuracy, decreasing = TRUE), ][c(1:25), ]
```


```{r}
ggplot(RF_status_classify_imp_25[c(1:25), ], aes(x = reorder(features, -MeanDecreaseAccuracy), MeanDecreaseAccuracy)) + geom_bar(stat="identity") +
  labs(title = "Most Important Features (Species)", subtitle = "Mean Decrease in Accuracy") +
  theme(plot.title = element_text(size=15, face="bold", hjust=1), plot.subtitle = element_text(size=10, hjust=0.5), 
        axis.title = element_text(size=10, face="bold"), axis.text = element_text(size=8, face="bold"), legend.position="bottom", 
        axis.text.x = element_text(angle = 90, hjust=0.5),
        axis.line = element_line(size = 0.3), legend.title = element_text(size=13, face="bold"), legend.text = element_text(size = 11))
ggsave(file = "figs/otu_table_filt_rare_removed_pct_mds.pdf", width = 10, height = 6, units = "in")
```



MeanDecreaseAccuracy: Measure of the extent to which a variable improves the accuracy of the forest in predicting the classification. Higher values mean that the variable improves prediction. In a rough sense, it can be interpreted as showing the amount of increase in classification accuracy that is provided by including the variable in the model

How many features (species) have positive Mean Decrease Accuracy?
```{r}
length(which(RF_state_classify_imp$MeanDecreaseAccuracy > 0))
important_features <- RF_state_classify_imp$features[RF_state_classify_imp$MeanDecreaseAccuracy > 0]
```

